---
title: "Course Project ML"
output:
  pdf_document: default
  html_document: default
---
Prediction Assignment Writeup. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.
## Loading and reading the data
```{r loading, echo=TRUE, cache= TRUE, warning=FALSE, include= TRUE}
library(RANN)
set.seed(1931)
traindata <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testdata<- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(traindata, destfile = "./pml-training.csv")
download.file(testdata, destfile = "./pml-testing.csv")

##read data replacing white spaces with "NA"
training <- read.table("pml-training.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
testing <- read.table("pml-testing.csv", header = TRUE, sep = ",", na.strings=c("","NA"))

##output data frame dimensions
dim(training) 
dim(testing) 
```

## Preprocess training set
Source data shows a a lot of missing values, por instance the variable max_roll_forearm:
```{r missing, echo=TRUE}
sum(is.na(training$max_roll_forearm))
```

We'll choose an arbitrary threshold (85%) and remove variables that have more than that percentage of NAs.

```{r clean, echo=TRUE}
##clean training set
cleantr <- training[, colMeans(is.na(training)) <= .15]
dim(cleantr)
```

We'll additionally remove predictors which are highly correlated (at or above Spearman's rho of 0.8), considerably reducing the number of predictors to 42.

```{r corr, echo=TRUE, warning=FALSE}
library(caret)
##clean training set
intermDF<- cleantr[,-c(1:6, 60)] ##remove non numeric variables
corintermDf<- cor(intermDF) #calculate correlations
highCor <- findCorrelation(corintermDf, cutoff=0.8) #set the cut off point at 0.8
CleanTraining <- cleantr[,-c(highCor)]
CleanTraining <- CleanTraining[,-c(1:5)]
dim(CleanTraining)
```

###Preprocess test set
The test set needs to be cleaned up so the variables match those left in the training set. We'll remove the variables from the testing set that were removed from the training set:

```{r cleante, echo=TRUE}
cols_to_keep <- intersect(colnames(CleanTraining),colnames(testing)) # store matching columns
CleanTraining <- CleanTraining[,cols_to_keep, drop=FALSE] # keep matching columns in training set
CleanTest <- testing[,cols_to_keep, drop=FALSE] # keep matching columns in test set
CleanTraining$classe<- training$classe # add a column with the outcome variable
CleanTest$classe<- testing$classe # add a column with the outcome variable
```

##Data splitting
We split our preprocessed training data 80/20 into training and test sets:

```{r split, echo=TRUE}
inTrain<- createDataPartition(y=CleanTraining$classe, p= 0.80, list=FALSE)
train_data<- CleanTraining[inTrain,]
test_data <-CleanTraining[-inTrain,]
```

##Fit a Random Forest

A random forest is a reasonable preliminary choice given the outcome is five categorical variables
```{r fit, echo=TRUE, cache=TRUE}
modelFit<- train(classe~.,method="rf", data= train_data, ntree=500)
modelFit$finalModel
```

We now see how it performs on the test set, i.e, its out of sample error. As shown below, the classification accuracy is very good. We'll therefore use it for our prediction exercise

```{r predict, echo=TRUE, cache=TRUE}
modelPredict<- predict(modelFit, newdata= test_data)
confuMat<- confusionMatrix(modelPredict, test_data$classe)
print(confuMat)
```

##Predict on the testing data frame
```{r predict2, echo=TRUE, cache=TRUE}
predictTest <- predict(modelFit, newdata= CleanTest)
print(predictTest)
```
